Interfacing logic programming and Coq

### 2020-10-26 papers/types2020/pbt.tex

\section{Property-based testing for Coq}

Simplest setting since we are only sending back to Coq a term (a
counterexample), not a general proof.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

>>>>

\section{Revisiting property-based testing for Coq}

In a previous paper~\cite{blanco19ppdp}, we presented a
proof-theoretical reconstruction of property-based testing (PBT) of
relational specifications, adopting techniques from foundational proof
certificates to account for several features of these testing
paradigm: from various \emph{generation} strategies, to
\emph{shrinking} and fault localization.

Given the connection that \textsf{coq-elpi} offers between logic
programming and the internals of Coq, it is natural to extend the FPC-driven logic programming interpreter of the previous section to perform PBT of \lstinline{Inductive} types.

While nothing prevents us from porting all the features we discussed in~\cite{blanco19ppdp}, for the sake of this paper we will implement only FPC corresponding to different flavors \emph{exhaustive} generation and their combination, as adopted, e.g., in SmallCheck~\cite{smallcheck} and  $\alpha$Check~\cite{cheney_momigliano_2017}.

Of course, \textsf{QuickChick}~\cite{QChick}
(\url{https://softwarefoundations.cis.upenn.edu/qc-current}) is a
sophisticated and far-reaching PBT tool for Coq, which is based on a
different perspective: being a clone of Haskell's QuickCheck, it
emphasizes testing executable (read \emph{decidable)} specifications
with random generators. While current research~\cite{LampropoulosPP18} aims to increase automation, it is fair to say testing, in particular of  relational specifications, is still very labor intensive. We do not intend to compete with  \textsf{QuickChick}, but as we shall see, as far as \lsti{Inductive} definitions that corresponds to pure Horn programs, we can test them immediately without any decidability proof or configuration of generators and shrinkers.

\subsection{PBT as proof reconstruction}
\label{ssec:pbt-lp}

\begin{metanote}
  I'm just saying the bare minimum -am
\end{metanote}

 If we view  a property as a logical formula
\(\forall x [(\tau(x)\wedge P(x)) \supset Q(x)]\) where $\tau$ is a
typing predicate and $P$ and $Q$ are two other predicates defined
using Horn clause specifications, providing a counter-example
consists of  negating the property, and \emph{searching} for a proof of
\[
  \exists x [(\tau(x)\wedge P(x)) \wedge \neg Q(x)]
  \tag{*}\label{eq:full}
\]
%\(\exists x [(\tau(x)\land P(x)) \land \neg Q(x)]\).

In our current setting $P$ and $Q$ will be \lsti{Inductive} propositions, while $\tau$   are honest-to-goodness datatypes. We will treat the two quite differently, in so far as elements of those types will be \emph{generated}, while predicates will only be \emph{checked}. This distinction plays also a part in interpreting the negation sign.
%
When using a focused proof system for logic extended with fixed
points~\cite{baelde12tocl}, negation corresponds to the usual intutionistic interpretation.  % proofs of formulas such as
% \[
%   \exists x [(\tau(x)\wedge P(x)) \wedge \neg Q(x)]
%   \tag{*}\label{eq:full}
% \]
% are a single \emph{bipole}: that is, when reading a proof bottom up, a
% positive phase is followed on all its premises by a single negative
% phase that completes the proof.%
% In particular, the positive phase corresponds to the \emph{generation} phase
% and the negative phase corresponds to the \emph{testing} phase.
% %

% Instead of giving a full focused proof system of a logic including
% fixed points (since, as we will argue, that proof system will not, in
% fact, be needed to account for PBT ), we offer the following analogy.
% %
% Suppose that we are given a finite search tree and we are asked to
% prove that there is a secret located in one of the nodes of that
% tree.
% %
% A proof that we have found that secret can be taken to be a
% description of the path to that node from the root of the tree: that
% path can be seen as the proof certificate for that claim.
% %
% On the other hand, a proof that no node contains the secret is a
% rather different thing: here, one expects to use a blind and
% exhaustive search (via, say, depth-first or breath-first search) and
% that the result of that search never discovers the secret.
% %
% A proof of this fact requires \emph{no} external information: instead it
% requires a lot of computation involved with exploring the tree until
% exhaustion.
% %
% This follows the familiar pattern where the positive (generate) phase
% requires external information while the negative (testing) phase
% requires none.
%
However, for the sake of PBT, we can identify a proof certificate for
$(\ref{eq:full})$ with a proof certificate for
\[
  \exists x [\tau(x)\wedge P(x)].
  \tag{**}\label{eq:short}
\]
and will resort to negation-as-failure to check that the conclusion does not hold. This also means that we will not be able to return a Coq proof terms for the refutation of our property --- and neither does \textsf{QuickChick}, which runs at the OCaml level --- although we can return the witness for the existential.
% Such a certificate would contain the witness (closed) term $t$ for the
% existential quantifier and sufficient information to confirm that
% $P(t)$ can be proved (a proof of a typing judgment such as $\tau(t)$
% is usually trivial).
% Since a proof certificate for the existence-of-a-counterexample formula
% $(\ref{eq:full})$ can be taken as a proof certificate of
% $(\ref{eq:short})$, then we only need to consider proof certificates
% for Horn clause programs.
%
%Since the logic of Horn clauses is rather simple,
% We illustrate next what such proofs and proof certificates look like
% for the rather simple logic of Horn clauses.

\begin{metanote}
  TODO next
  \begin{itemize}
  \item The elpi dep-pbt tactic
  \item the simple reverse example
  \item a bit of details about the \texttt{aexp} example
  \end{itemize}
\end{metanote}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

% LocalWords:  Bedwyr bipole elpi QuickCheck OCaml TODO
